Developing Observational Health Research Capacity in the Julia Ecosystem

### 2023-12-23 paper.tex
Added additional todos and experimented with code listings

One of the premier tools within the OHDSI HADES ecosystem is the ATLAS Shiny application which was developed to develop patient phenotype definitions within observational health research contexts (amongst other use cases).
These phenotype definitions have varying levels of complexity where they can be very simple (such as finding all patients with one health condition or diagnosis) or grow to be extremely complex (defining controlled, uncontrolled, and indeterminate patient cohorts based on medications, lab results, etc.)
% TODO: Cite schema of the definition from GitHub
The computable phenotype definitions that ATLAS generates based on phenotype definitions are in a variety of serialization formats such as JSON or prepared SQL expressions available in multiple SQL flavors.

% TODO: Cite package
% TODO: Reframe this to more say how ATLAS can be used to build cohorts
% TODO: Explain how this package to accomplish similar tasks
The Julia package, OHDSICohortExpressions.jl, exists specifically to ingest these serialized expressions of computable phenotype definitions and create patient cohorts that match the given criterion in the definition.
% TODO: Change this list into an enumeration
How this tool works is to 1) read and verify the serialized computable phenotype definition 2) reinterpret the serialization into the FunSQL.jl Domain Specific Language 3) return a prepared SQL statement against a given SQL database that queries an OMOP CDM database for the desired patient population(s) of interest.
Once these prepared statements are made, researchers can execute the generated SQL against their OMOP CDM instance to generate cohorts for future analyses.

@@

Although OHDSICohortExpressions.jl is sufficient to translate existing phenotype definitions generated in ATLAS to executable queries, oftentimes, there is a tricky problem of how to best iterate on phenotype definitions. \cite{zelkoDevelopingRobustComputable2023}
Often, one has to take an existing definition and either revise to a more narrow cohort or write manual code to build more niche cohorts after an initial definition is created.
This refactoring of a phenotype definition can lead to conflating ideas within conversation between expert clinical reviewers, analysts, and health researchers.
This broadly can be the notions of what we are exploring (the phenotype definition; often the responsibility of a clinical reviewer), what is possible to compute upon (the computable phenotype definition; often the responsibility of an analyst), and the goals of the overall study.
For situations such as these, the tool OMOPCDMCohortCreator.jl was created.

OMOPCDMCohortCreator.jl allows one to iteratively build a patient population quickly and easily.
One can use patient pools developed from pre-existing cohorts defined off of phenotype definitions as a starting point in exploring patients or create new patient populations quickly without the need for a well-defined phenotype definition.
This can enable analysts to more quickly understand what is inside their data assets, do "quick" analysis of a particular smaller research question, and enable a more distinct separation of work between a clinical expert reviewer and analyst teams.
It is highly recommended to still create phenotype definitions about particular populations one explores.

% TODO: Mention capr here somehow

@@

One of the biggest styming factors within not only the Julia ecosystem, but observational health research in general has been the difficulty of working with different databases.
Although the OMOP CDM has been great for standardizing \textit{how} real world data is handled, there is a lack of consensus on \textit{what} is needed to handle these data assets.
For a variety of reasons, different groups will choose different database architectures (perhaps PostgreSQL for improved permission support, DuckDB for performance, or even SQLite for ease of use) but all are subtly different enough to create headaches in seamlessly working across different research sites.

The way that the Julia ecosystem handles this is via the creation of a common database interface called the DBInterface.
Specific Julia database packages can implement dispatches for this interace allowing users to have a much better experience working across databases.
In spite of the such benefits, for some users (and especially within the context of observational health studies), it can beneficial to have a more prescriptive interface where one just needs to pass perhaps a password, username, database name, and schema.
For that reason, we DBConnector.jl was created.

Although this is not an explicit JuliaHealth package, it was a pain point that was experienced throughout the process of observational health research across multiple partner sites.
The promise of DBConnector.jl is that with one line command, users can connect simply, and in a common (albeit opinionated) manner, to databases.
DBConnector unifies Julia database packages that implement the DBInterface to achieve this level of simplified connection making.

Finally, if someone needs more fine-grained control over a particular database connection, they can choose to eschew this package.
Instead, they can switch to the specific Julia database package that would give them the additional control.
Although they'll lose the simplicity of this package, core required functionality is not lost when working with databases.

DBConnector.jl was inspired by the package \textit{DatabaseConnector} (Link: https://github.com/OHDSI/DatabaseConnector)

% TODO: Add database connector example
% TODO: Reference database connector example

\subsection{Ecosystem Interoperability}

@@

With the JuliaHealth tools that have been created, we can now demonstrate an observational health study workflow that would be used in practice.
For this paper, we used two datasets: a synthetic patient dataset called \textit{Eunomia} and the MIMIC III patient dataset which had been converted into the OMOP CDM to illustrate the workflow provided by the Observational Health sub-ecosystem.
Reproducing the workflow using the \textit{Eunomia} dataset is made readily available at the repository here: \cite{schuemieEunomia2023}

% TODO: Add tutorial to separate repository

@@

\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
import DBConnector: 
  DBConnection
import HealthSampleData: 
  Eunomia
import OMOPCDMCohortCreator as OCC

@@

occ.GenerateTables(conn)
\end{minted}

>>>>

% TODO: Add a paragraph here about what is a phenotype definition
% TODO: Cite my paper here

% TODO: Segue this better from phenotype definitions
One of the premier tools within the OHDSI HADES ecosystem is the ATLAS Shiny application which was developed to develop patient phenotype definitions within observational health research contexts (amongst other use cases).
These phenotype definitions have varying levels of complexity where they can be very simple (such as finding all patients with one health condition or diagnosis) or grow to be extremely complex (defining controlled, uncontrolled, and indeterminate patient cohorts based on medications, lab results, etc.)
% TODO: Cite schema of the definition from GitHub
% EXPLAIN: Better introduce what SQL is
The computable phenotype definitions that ATLAS generates based on phenotype definitions are in a variety of serialization formats such as JSON or prepared SQL expressions available in multiple SQL flavors.

% TODO: Cite package
% TODO: Reframe this to more say how ATLAS can be used to build cohorts
% TODO: Explain how this package to accomplish similar tasks
The Julia package, OHDSICohortExpressions.jl, exists specifically to ingest these serialized expressions of computable phenotype definitions and create patient cohorts that match the given criterion in the definition.
% TODO: Change this list into an enumeration
How this tool works is to 1) read and verify the serialized computable phenotype definition 2) reinterpret the serialization into the FunSQL.jl Domain Specific Language 3) return a prepared SQL statement against a given SQL database that queries an OMOP CDM database for the desired patient population(s) of interest.
% EXPLAIN: What is a prepared statement?
% EXPLAIN: What is a cohort in this context?
Once these prepared statements are made, researchers can execute the generated SQL against their OMOP CDM instance to generate cohorts for future analyses.

@@

% TODO: Reword this sentence
Although OHDSICohortExpressions.jl is sufficient to translate existing phenotype definitions generated in ATLAS to executable queries, oftentimes, there is a tricky problem of how to best iterate on phenotype definitions. \cite{zelkoDevelopingRobustComputable2023}
% EXPLAIN: Better contextualize why this is important -- why not just create a new phenotype definition?
Often, one has to take an existing definition and either revise to a more narrow cohort or write manual code to build more niche cohorts after an initial definition is created.
% EXPLAIN: How so?
% TODO: Review this sentence; maybe cut it
This refactoring of a phenotype definition can lead to conflating ideas within conversation between expert clinical reviewers, analysts, and health researchers.
% TODO: Move this to the previous section on phenotype definitions
This broadly can be the notions of what we are exploring (the phenotype definition; often the responsibility of a clinical reviewer), what is possible to compute upon (the computable phenotype definition; often the responsibility of an analyst), and the goals of the overall study.

% TODO: Present this more as a general introduction to what tools are here for this
% TODO: Talk about composition with DataFrames or Tidier
OMOPCDMCohortCreator.jl allows one to iteratively build a patient population quickly and easily.
One can use patient pools developed from pre-existing cohorts defined off of phenotype definitions as a starting point in exploring patients or create new patient populations quickly without the need for a well-defined phenotype definition.
% TODO: Mention the ability to satisfy research requests
This can enable analysts to more quickly understand what is inside their data assets, do "quick" analysis of a particular smaller research question, and enable a more distinct separation of work between a clinical expert reviewer and analyst teams.

% TODO: Mention capr

@@

% TODO: Mention JuliaCon Birds of a Feather discussion
One of the biggest styming factors within not only the Julia ecosystem, but observational health research in general has been the difficulty of working with different databases.
Although the OMOP CDM has been great for standardizing \textit{how} real world data is handled, there is a lack of consensus on \textit{what} is needed to handle these data assets.
% TODO: Make this less verbose
For a variety of reasons, different groups will choose different database architectures (perhaps PostgreSQL for improved permission support, DuckDB for performance, or even SQLite for ease of use) but all are subtly different enough to create headaches in seamlessly working across different research sites.

% TODO: Change the phrasing of this
The way that the Julia ecosystem handles this is via the creation of a common database interface called the DBInterface.
% EXPLAIN: What is dispatch
Specific Julia database packages can implement dispatches for this interace allowing users to have a much better experience working across databases.
In spite of such benefits, for some users (and especially within the context of observational health studies), it can beneficial to have a more prescriptive interface where one just needs to pass perhaps a password, username, database name, and schema.

% TODO: Better introduce this
For that reason, DBConnector.jl was created.
% EXPLAIN: Contextualize this a bit better
Although this is not an explicit JuliaHealth package, it was a pain point that was experienced throughout the process of observational health research across multiple partner sites.
The promise of DBConnector.jl is that with one line command, users can connect simply, and in a common (albeit opinionated) manner, to databases.
DBConnector unifies Julia database packages that implement the DBInterface to achieve this level of simplified connection making.
DBConnector.jl was inspired by the package \textit{DatabaseConnector} (Link: https://github.com/OHDSI/DatabaseConnector)

% TODO: Combine with the previous paragraph and state that there are two approaches
% TODO: Make explicit DBConnector's path
% TODO: Make explicit the path of choosing general packages
Finally, if someone needs more fine-grained control over a particular database connection, they can choose to eschew this package.
Instead, they can switch to the specific Julia database package that would give them the additional control.
Although they'll lose the simplicity of this package, core required functionality is not lost when working with databases.

% TODO: Review if this section is actually needed
\subsection{Ecosystem Interoperability}

@@

% TODO: Reference code repository
Using the JuliaHealth tools developed, we enumerate an example observational health study workflow that would be used in practice.\footnotemark
For this paper, we used two datasets: a synthetic patient dataset called \textit{Eunomia} and the MIMIC III patient dataset which had been converted into the OMOP CDM to illustrate the workflow provided by the Observational Health sub-ecosystem.
Reproducing the workflow using the \textit{Eunomia} dataset is made readily available at the repository here: \cite{schuemieEunomia2023}

% TODO: Add link to code repository
\footnotetext{The code snippets presented in the body of the paper may omit certain intermediate steps. 
Please review the associated code repository for a full working tutorial.}

@@

\begin{listing}[!ht]
\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
import DBConnector: 
  DBConnection
import HealthSampleData: 
  Eunomia
import OMOPCDMCohortCreator as OCC

@@

occ.GenerateTables(conn)
\end{minted}
\caption{This illustrates how to connect to the Eunomia database}
\label{listing:connection}
\end{listing}
